One-The-Fly Decision Procedures For GKAT

### 2024-12-02 main.tex
symbolic algorithm

\begin{algorithm}
    \caption{Check whether a state \(s\) is dead}\label{alg:check-dead-main}
    \begin{algorithmic}
        \Function{isDeadLoop}{$s ∈ S$, explored}
        \If {\(s ∈\) explored} {\Return explored} 
        \Else { 
            \For{\(α ∈ \At\)}{}{
                \Match{\(δ_{S}(s, α)\)}
                \Case{\(\accept\)} {\Return none} \Comment{\(s\) transition to accept}\EndCase
                \Case{($s', p$)}{ 
                    \If{\Call{IsDead}{s'} = none} {\Return none} \Comment{\(s\) transitions to a live state \(s'\)}
                    \Else { explored \(←\) (explored \(∪\) \Call{isDeadLoop}{$s'$, explored}) } \EndIf
                } \EndCase
                \EndMatch
            }\EndFor}
        \EndIf
        \State {\Return explored}    
        \EndFunction
    \end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
    \caption{On-the-fly bisimulation algorithm}\label{alg:bisim}
    \begin{algorithmic}
        \Function{equiv}{$s ∈ S, t ∈ T$}
        \If {\Call{eq}{$s, t$}} {\Return true}
        \ElsIf {\(s ∈\) deadStates} {\Return \Call{isDead}{$t$}} 
        \ElsIf {\(t ∈\) deadStates} {\Return \Call{isDead}{$s$}} 
        \Else 
        \For{\(α ∈ \At\)}{}{
            \Comment{Inductive construction,~\cref{thm:inductive-construction}}
            \Match{$δ_{S}(s, α), δ_{T}(t, α)$}
            \Case{\(\accept, \accept\)} {\Skip} \EndCase
            \Case{\(\reject, \reject\)} {\Skip} \EndCase
            \Case{\(\reject, (t', q)\)} {\Call{isDead}{$t'$}} \EndCase
            \Case{\((s', p), \reject\)} {\Call{isDead}{$s'$}} \EndCase
            \Case{\((s', p), (t', q)\)} {
                \If {\(p = q\)} {\Call{union}{$s, t$}; \Call{equiv}{$s, t$}} 
                \ElsIf {\Call{isDead}{$s$} and \Call{isDead}{$s$}} {\Skip}
                \Else { \Return false }
                \EndIf
            } \EndCase
            \Default {\Return false} \Comment{the results format does not match} \EndDefault
            \EndMatch
        }\EndFor
        \EndIf
        \Return true \Comment{no mismatch found}
        \EndFunction
    \end{algorithmic}
\end{algorithm}

The soundness and completeness of~\cref{alg:bisim} can be observed by the fact that \emph{when the algorithm terminate}, the algorithm returns true if and only if there exists a bisimulation between \(⟨s⟩\) and \(⟨t⟩\) s.t. \(s ∼ t\), which is then logically equivalent to trace equivalence.
Such equivalence is a direct consequence of~\cref{thm:bisim-one-dead,thm:inductive-construction}.

@@

>>>>

TODO: we should merge the two so that it is easier to 
\begin{algorithm}
    \caption{Check whether a state \(s\) is dead}\label{alg:check-dead-main}
    \begin{algorithmic}
        \Function{isDeadLoop}{$s ∈ S$, explored}
        \If {\(s ∈\) explored} {\Return explored} 
        \Else { 
            \For{\(α ∈ \At\)}{}{
                \Match{\(δ_{S}(s, α)\)}
                \Case{\(\accept\)} {\Return none} \Comment{\(s\) transition to accept}
                \EndCase
                \Case{\(\reject\)} {\texttt{continue}}
                \Comment{skip if \(s\) transition to reject}
                \EndCase
                \Case{($s', p$)}{ 
                    \If{\Call{IsDeadLoop}{$s'$} = none} {\Return none} \Comment{\(s\) transitions to a live state \(s'\)}
                    \Else { explored \(←\) (explored \(∪\) \Call{isDeadLoop}{$s'$, explored}) } \EndIf
                } \EndCase
                \EndMatch
            }\EndFor}
        \EndIf
        \State {\Return explored}    
        \EndFunction
    \end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
    \caption{On-the-fly bisimulation algorithm}\label{alg:bisim}
    \begin{algorithmic}
        \Function{equiv}{$s ∈ S, t ∈ T$}
        \If {\Call{eq}{$s, t$}} {\Return true}
        \ElsIf {\(s ∈\) deadStates\(_S\)} {\Return \Call{isDead\(_T\)}{$t$}} 
        \ElsIf {\(t ∈\) deadStates\(_T\)} {\Return \Call{isDead\(_S\)}{$s$}} 
        \Else 
        \For{\(α ∈ \At\)}{}{
            \Comment{Inductive construction,~\cref{thm:inductive-construction}}
            \Match{$δ_{S}(s, α), δ_{T}(t, α)$}
            \Case{\(\accept, \accept\)} {\Skip} \EndCase
            \Case{\(\reject, \reject\)} {\Skip} \EndCase
            \Case{\(\reject, (t', q)\)} {\Call{isDead}{$t'$}} \EndCase
            \Case{\((s', p), \reject\)} {\Call{isDead}{$s'$}} \EndCase
            \Case{\((s', p), (t', q)\)} {
                \If {\(p = q\)} {\Call{union}{$s, t$}; \Call{equiv}{$s, t$}} 
                \ElsIf {\Call{isDead}{$s$} and \Call{isDead}{$s$}} {\Skip}
                \Else { \Return false }
                \EndIf
            } \EndCase
            \Default {\Return false} \Comment{the results format does not match} \EndDefault
            \EndMatch
        }\EndFor
        \EndIf
        \Return true \Comment{no mismatch found}
        \EndFunction
    \end{algorithmic}
\end{algorithm}

Because the dead state detection algorithm is coalgebra-specific, we use a subscript on ``deadStates'' and ``\textsc{IsDead}'' to indicate the coalgebra. 
The soundness and completeness of~\cref{alg:bisim} can be observed by the fact that \emph{when the algorithm terminate}, the algorithm returns true if and only if there exists a bisimulation between \(⟨s⟩\) and \(⟨t⟩\) s.t. \(s ∼ t\), which is then logically equivalent to trace equivalence.
Such equivalence is a direct consequence of~\cref{thm:bisim-one-dead,thm:inductive-construction}.

@@

Then for the algorithm, we can just recursively check all the conditions in~\cref{thm:symb-inductive-construction}.

\begin{algorithm}
    \caption{On-the-fly bisimulation algorithm}\label{alg:bisim}
    \begin{algorithmic}
        \Function{equiv}{$s ∈ S, t ∈ T$}
        \If {\Call{eq}{$s, t$}} {\Return true}
        \ElsIf {\(s ∈\) deadStates\(_S\)} {\Return \Call{isDead\(_T\)}{$t$}} 
        \ElsIf {\(t ∈\) deadStates\(_T\)} {\Return \Call{isDead\(_S\)}{$s$}} 
        \Else {}
        \Return {
            \Comment{conditions of ~\cref{thm:symb-inductive-construction}}
            \\\vspace{5px}
            \(\qquad
            \begin{aligned}
                & ϵ̂_S(s) ≡ ϵ̂_T(t) \mathrel{\&\!\&} \\  
                & \forall (b, s', p) ∈ δ̂_S(s), (c, t', q) ∈ δ̂_T(t), (b ∧ c) ≢ ⊥ ⟹ 
                \begin{cases}
                    \text{\Call{IsDead$_S$}{$s$}} ∧ \text{\Call{IsDead$_T$}{$t$}} & \text{if \(p ≠ q\)} \\
                    \text{\Call{Union}{$s$, $t$}}; \text{\Call{Equiv}{$s', t'$}} & \text{if \(p = q\)}
                \end{cases} \mathrel{\&\!\&}\\
                & \forall (b, s', p) ∈ δ̂_S(s), c ∈ ρ̂_T(t), (b ∧ c) ≢ ⊥ ⟹ \text{\Call{IsDead$_S$}{$s'$}} \mathrel{\&\!\&}\\
                & \forall b ∈ ρ̂_S(s), (c, t', q) ∈ ρ̂_T(t), (b ∧ c) ≢ ⊥ ⟹ \text{\Call{IsDead$_T$}{$t'$}}
            \end{aligned}\)
        }
        \EndIf
        \EndFunction
    \end{algorithmic}
\end{algorithm}

Inspired by the syntax of Ocaml, the \(\mathrel{\&\!\&}\) is the logical-and operator on the language level, specifying that all four conditions in the return statements must be satisfied to return true. 
Notice just like the non-symbolic case, this algorithm can be modified to perform symbolic bisimulation of (non-normalized) GKAT automaton, which coincides with infinite trace equivalence~\cite{schmid_GuardedKleeneAlgebra_2021}, by letting \textsc{IsDead} always return false and keep deadStates empty.

\section{Construction of Symbolic GKAT Automata}

We will go through two ways of constructing symbolic GKAT automata, one is Thompson's construction, the other is derivative.
However, in order to exhibit the relation between these two construction, specifically, the derivative construction always produces a smaller automaton than the Thompson's construction, we opt to use to produce "weak" symbolic GKAT automata: symbolic GKAT automata where the transition no longer need to perform an action.

\subsection{Weak Symbolic GKAT Automata And Tightening}

Weak Symbolic GKAT automata is a symbolic GKAT automata that allow internal transitions via a boolean expression:
\[⟨S, 
ϵ̂: S → \Bool, 
δ̂: S → (\Bool × S) + (\Bool × S × K)⟩.\]
We also require the deterministic criteria, all the boolean transitions of \(s\), i.e. all the boolean expressions in the following set:
\[\{ϵ̂(s)\} ∪ 
\{b ∣ ∃ s' ∈ S, (b, s') ∈ δ̂(s)\} ∪ 
\{b ∣ ∃ s' ∈ S, p ∈ K, (b, s', p) ∈ δ̂(s)\},\]
are disjoint.
Notice that our notion of
